

# Random Sampling

```{r}
library(IntroStats)
```

## Random Number Generators

The `runif()` function generates random numbers from a **uniform distribution** over the interval [0, 1] by default.

```{r}
set.seed(123)
uniform_sample <- runif(100)
hist(uniform_sample, col = "skyblue", main = "Uniform(0,1) Distribution", xlab = "Value")
```

You can also generate numbers from any interval [a,b]:

```{r}
runif(5, min = 10, max = 20)

```

### Random Numbers Sampling Using a Seed

To ensure reproducibility, use set.seed() before generating random numbers. This ensures the same output each time.

```{r}
set.seed(2025)
runif(3)
```

### Random Numbers from a Normal Distribution

```{r}
set.seed(42)
normal_sample <- rnorm(1000, mean = 0, sd = 1)
hist(normal_sample, col = "salmon", main = "Standard Normal Distribution", xlab = "Value")

```

### Random Numbers from a Binomial Distribution

```{r}
set.seed(10)
binomial_sample <- rbinom(1000, size = 10, prob = 0.5)
hist(binomial_sample, col = "lightgreen", main = "Binomial(n=10, p=0.5)", xlab = "Successes")

```

### Random Numbers from a Poisson Distribution

Use rpois(n, lambda) where lambda is the average rate of occurrence.

```{r}
poisson_sample <- rpois(1000, lambda = 5)
hist(poisson_sample, col = "lightblue", main = "Poisson(λ = 5)", xlab = "Count")

```

### Random Numbers from a t Distribution

Use rt(n, df) for random values from a Student's t distribution.

```{r}
t_sample <- rt(1000, df = 10)
hist(t_sample, col = "orange", main = "t-distribution (df=10)", xlab = "Value")
```

### Random Numbers from an F Distribution

Use rf(n, df1, df2) to generate values from an F-distribution.

```{r}
f_sample <- rf(1000, df1 = 5, df2 = 10)
hist(f_sample, col = "plum", main = "F(5,10) Distribution", xlab = "Value", xlim = c(0,5))
```

### Random Numbers from a Chi-square Distribution

Use rchisq(n, df) for Chi-square distributed values.

```{r}
chi_sample <- rchisq(1000, df = 4)
hist(chi_sample, col = "khaki", main = "Chi-square(df=4)", xlab = "Value")
```

### Examples

**Example 1**

```{r}
#Sample from population
officials = c("G", "L", "S", "A", "T")

#Draw a sample of 2 officials without replacement
sample(officials, size = 2, replace = FALSE)

#Draw a sample of 4 officials without replacement
sample(officials, size = 4, replace = FALSE)

#Draw a sample of 4 officials with replacement
sample(officials, size = 4, replace = TRUE)

```
**Example2**

```{r}

#Stratified sampling proportional allocation

stratum1 = 1:400 #a set of integers from 1 to 400
stratum2 = 401:1000 #a set of integers from 401 to 1000
stratum3 = 1001:1800
stratum4 = 1801:2000

#Total sample size
N = 10

#size by each stratum

n1 = N * 400 / 2000
n2 = N * 600 / 2000
n3 = N * 800 / 2000
n4 = N * 200 / 2000

#sample by each stratum
s1 = sample(stratum1, size=n1, replace=FALSE)
s2 = sample(stratum2, size=n2, replace=FALSE)
s3 = sample(stratum3, size=n3, replace=FALSE)
s4 = sample(stratum4, size=n4, replace=FALSE)

#combine stratum samples together
s = c(s1, s2, s3, s4)
s1
s2
s3
s4

s
```

**Example3**

```{r}

#Stratified sampling proportional allocation

stratum1 = 1:25 #a set of integers from 1 to 25
stratum2 = 26:200 #a set of integers from 26 to 200
stratum3 = 201:250

#Total sample size
N = 20

#size by each stratum

n1 = N * 25 / 250
n2 = N * 175 / 250
n3 = N * 50 / 250

#sample by each stratum
s1 = sample(stratum1, size=n1, replace=FALSE)
s2 = sample(stratum2, size=n2, replace=FALSE)
s3 = sample(stratum3, size=n3, replace=FALSE)

#combine stratum samples together
s = c(s1, s2, s3)
s1
s2
s3

s
```

## Random Sampling from a Sample

Random sampling is often used to estimate population quantities or build simulations.

### Bootstrap Sampling with Replacement

Bootstrap is a resampling technique that involves repeatedly drawing samples with replacement from the original dataset. This method allows us to estimate the sampling distribution of a statistic, such as the mean.

```{r}
data <- c(5, 7, 8, 10, 13)
set.seed(1)
bootstrap_sample <- sample(data, size = 5, replace = TRUE)
bootstrap_sample
```
In the code below, we generate a single bootstrap sample and then repeat this process 1,000 times to compute the mean of each sample. The histogram displays the distribution of these bootstrap means.

```{r}
data <- c(5, 7, 8, 10, 13)
set.seed(1)
B <- 1000
means <- replicate(B, mean(sample(data, size = 5, replace = TRUE)))
length(means) 
head(means)   

hist(means,
     breaks = 30,
     col = "gray",
     main = "Bootstrap Sampling Distribution of Mean",
     xlab = "Sample Mean")
```


### Random Sampling without Replacement
Use sample() without the replace=TRUE argument to select unique items.

```{r}

set.seed(123)
sample(data, size = 3, replace = FALSE)
```


## Stratified sampling with proportional allocation

The following R function is to sample data with proportional allocation
```{r}
strat.sample = function(x, strata, size, proportional = TRUE){
  #x: an array that represents the entire population
  #strata: stratum indicator
  #size: sample size
  #proportional: TRUE or FALSE. If true, the sample is proportional to the strata size.
  
  n = length(x)
  u.strata = unique(strata)
  n.strata = length(u.strata)
  m = rep(NA, n.strata) #size of each stratum
  
  for (i in 1:n.strata){m[i] = length(strata[strata == u.strata[i]])}
  ni = round(size*m/n)
  if (sum(ni) > size) {ni[n.strata] = ni[n.strata] - (sum(ni) - size)} else if (sum(ni) < size) {ni[n.strata] = ni[n.strata] + (size-sum(ni))}
  
  if(proportional){
      s = NULL
      for (i in 1:n.strata){
        s = c(s, sample(x[strata == strata[i]], size=ni[i], replace=FALSE))
      }
  } else {
    s= sample(x, size=size, replace=FALSE)
  }
  return(s)
}

strat.sample(x=rnorm(100), strata=c(rep(1, 20), rep(2, 40), rep(3, 40)), size=10, proportional = TRUE)

strat.sample(x=rnorm(100), strata=c(rep(1, 20), rep(2, 40), rep(3, 40)), size=12, proportional = TRUE)
```





