
# Distribution Shapes

```{r}
library(IntroStats)
```

## Distribution of a dataset

The **distribution** of a dataset describes the values and how often they occur. It can be shown as:
- Frequency or relative frequency tables
- Histograms for continuous data
- Density curves or probability distribution functions
- Cumulative distribution functions (CDF)
Understanding the distribution helps us see the center (mean/median), spread (variance/IQR), and shape (modality, symmetry, skewness, tails).

### Example

A class of 12 students scored:
`58, 60, 61, 65, 66, 68, 70, 72, 75, 78, 80, 85`. Create a frequency table and histogram, and mark the mean and median.

**Solution:**
```{r}
scores <- c(58,60,61,65,66,68,70,72,75,78,80,85)
tab <- table(cut(scores, breaks = seq(55, 90, by=5), right = FALSE))
tab
mean(scores); median(scores)
hist(scores, breaks = seq(55, 90, by=5), prob = TRUE, main = "Scores Histogram",
     xlab = "Score")
lines(density(scores), lwd=2)
abline(v = mean(scores), lwd=2, lty=2)
abline(v = median(scores), lwd=2, lty=3)
legend("topright", legend=c("Mean","Median"), lty=c(2,3), bty="n")
```
The mean and median are close, showing a relatively symmetric distribution.

### Example

```{r}
height = rnorm(1000, mean=64, sd=3)
hist(height, prob = TRUE, col="purple", breaks = 20)
lines(density(height), col = 4, lwd = 2)
```


## Modality

Examples of (a) unimodal, (b) bimodal, and (c) multimodal distributions, defined based on the number of peaks.

```{r}
#Unimodal
draw.multi.modes(peaks.loc = c(5),  width=c(3), 
                 xlim=c(-5, 15), ylim=c(0, 0.42), 
                 main="Peak: 5, width: 3")

draw.multi.modes(peaks.loc = c(5),  width=c(2), 
                 xlim=c(-5, 15), ylim=c(0, 0.42), 
                 main="Peak: 5, width: 2")

draw.multi.modes(peaks.loc = c(5),  width=c(1), 
                 xlim=c(-5, 15), ylim=c(0, 0.42), 
                 main="Peak: 5, width: 1")

```

```{r}
#Bimodal
draw.multi.modes(peaks.loc = c(5, 15),  width=c(1,1),
                 xlim=c(-5, 25), ylim=c(0, 0.25), 
                 main="Peak: 5 15, width: 1 1")
draw.multi.modes(peaks.loc = c(5, 15),  width=c(1,2),
                 xlim=c(-5, 25), ylim=c(0, 0.25), 
                 main="Peak: 5 15, width: 1 2")
draw.multi.modes(peaks.loc = c(5, 15),  width=c(1,3),
                 xlim=c(-5, 25), ylim=c(0, 0.25), 
                 main="Peak: 5 15, width: 1 3")
draw.multi.modes(peaks.loc = c(5, 15),  width=c(2,2),
                 xlim=c(-5, 25), ylim=c(0, 0.25), 
                 main="Peak: 5 15, width: 2 2")
draw.multi.modes(peaks.loc = c(5, 15),  width=c(3,3),
                 xlim=c(-5, 25), ylim=c(0, 0.42), 
                 main="Peak: 5 15, width: 3 3")

```

```{r}
#3-modal
draw.multi.modes(peaks.loc = c(5, 15, 25),  width=c(3,3,3),
                 xlim=c(-5, 35), ylim=c(0, 0.15), 
                 main="Peak: 5 15 25, width: 3 3 3")
draw.multi.modes(peaks.loc = c(5, 15, 25),  width=c(1,2,3),
                 xlim=c(-5, 35), ylim=c(0, 0.15), 
                 main="Peak: 5 15 25, width: 1 2 3")
draw.multi.modes(peaks.loc = c(5, 15, 25),  width=c(3,2,3),
                 xlim=c(-5, 35), ylim=c(0, 0.15), 
                 main="Peak: 5 15 25, width: 3 2 3")
```

## Symmetry

Examples of symmetric distributions: (a) bell shaped, (b) triangular, and (c) uniform

```{r}
#bell shaped
draw.bell.shape(peak.loc=10, width=2, main="Bell Shaped Distribution")

```


```{r}
#Triangular shaped
draw.triangle.shape(peak.loc=10, slope=0.25, n=20,
                    main="Triangle Shaped Distribution",
                    ylim=c(0, 0.5), xlim=c(7, 13))

draw.triangle.shape(peak.loc=10, slope=1, n=20,
                    main="Triangle Shaped Distribution",
                    ylim=c(0, 0.5), xlim=c(7, 13))

draw.triangle.shape(peak.loc=10, slope=4, n=20,
                    main="Triangle Shaped Distribution",
                    ylim=c(0, 0.5), xlim=c(7, 13))

```



```{r}
#Uniform shaped
draw.uniform.shape(range = c(0, 1), main="Uniform Distribution")

```


### Skewness

Skewed distributions: (a) right skewed (b) left skewed

```{r}
draw.skewed.shape(peak.loc=10, width=1, skew = "right", 
                  main="Right Skewed Distribution",
                  xlim=c(0, 50), ylim=c(0, 0.22))

draw.skewed.shape(peak.loc=10, width=2, skew = "right", 
                  main="Right Skewed Distribution",
                  xlim=c(0, 50), ylim=c(0, 0.22))

draw.skewed.shape(peak.loc=10, width=2, skew = "left", 
                  main="Left Skewed Distribution",
                  xlim=c(-20, 20), ylim=c(0, 0.17))

draw.skewed.shape(peak.loc=10, width=3, skew = "left", 
                  main="Left Skewed Distribution",
                  xlim=c(-20, 20), ylim=c(0, 0.17))

draw.skewed.shape(peak.loc=10, width=2, skew = "no", 
                  main="Not Skewed (symmetric) Distribution",
                  xlim=c(0, 20), ylim=c(0, 0.22))

draw.skewed.shape(peak.loc=10, width=3, skew = "no", 
                  main="Not Skew (symmetric) Distribution",
                  xlim=c(0, 20), ylim=c(0, 0.22))

```

Another type of skewed distribution is reverse-J shaped.

```{r}
draw.reverse.Jshape (median.loc=10)

```



